---
title: "load_and_analyze"
output: html_document
---

#grey areas are code blocks
#green play button to run code block

#20 obs of 15 vars = 20 rows of 15 columns

## Setup

```{r}
#
# Install necessary libraries if not installed
#

# if tidyverse not installed, install
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

#
# Load libraries
#

library(tidyverse)

```

## Load Data

```{r}
#
# Load data
#

bp_fy2020 <- read_csv("bp_fy_2020.csv")

```
## Get a Sense of the Data
```{r}
#
# Get a sense of it
#

# Examine columns, data types, select values
bp_fy2020 |>
  glimpse()

# View it in interactive viewer
View(bp_fy2020)

```


## Ask Questions

### Arrange, Select

Q: Which sector has the highest number of apprehensions in FY 2020?
Method: Arrange, Select
A: 
```{r}

#piping is taking the thing on the left and putting it in the function on the right
#piping can chain functions
#select in "select(sector, yearly_total)" means get rid of all columns except sector and yearly total 
#arrange is using desc for descending order
#default order is lowest to highest

bp_fy2020 |>
  arrange(desc(yearly_total)) |> 
  select(sector, yearly_total)

```

Q: Which sector has the lowest number of apprehensions in FY 2020?
Method: Arrange, Select
A: 

```{r}

bp_fy2020 |>
  arrange(yearly_total) |> 
  select(sector, yearly_total)

```

### Filter

Q: How many apprehensions did Yuma have in FY 2020?
Method: Select, Filter
A: 

```{r}

#filter lets you select a column, then the select narrows down the columns to yearly_total

bp_fy2020 |>
  filter(sector == "Yuma") |> 
  select(sector, yearly_total)

```

### Summarize
Q: How many apprehensions were there in FY 2020 across all sectors?
Method: Summarize, Sum
A: 

```{r}

#summarize function creates a new column called total_apprehensions and puts in the sum of the yearly_total column

bp_fy2020 |>
  summarize(total_apprehensions = sum(yearly_total, na.rm = TRUE))

```

Q: What was the average number of apprehensions in each sector in FY 2020?
Method: Summarize, Mean
A: 

```{r}

#same as above but using key word average

bp_fy2020 |>
  summarize(average_apprehensions = mean(yearly_total, na.rm = TRUE))

```

### Creating New Columns 
Q: How many apprehensions were there in Miami in October, November and December of FY2020?
Method: Mutate, Filter
A: 
```{r}

#first pipe is creating a new column that adds up those columns and puts it in fy_2020_Q1
#next pipe filters to just Miami
#next pipe selects just columns we care about

bp_fy2020 |>
  mutate(fy_2020_Q1 = october + november + december) |>
  filter(sector == "Miami") |> 
  select(sector, fy_2020_Q1)


```

Q: What percentage of apprehensions in Miami were in October, November and December in FY2020?
Method: Mutate, Filter, Select 
A: 
```{r}

#first three columns do the same as above
#mutate key word to do arithmetic

bp_fy2020 |>
  mutate(fy_2020_Q1 = october + november + december) |>
  filter(sector == "Miami") |> 
  select(sector, fy_2020_Q1, yearly_total) |>
  mutate(percentage = fy_2020_Q1 / yearly_total * 100)


```
### Grouping and summarizing

Q: What was the total number of apprehensions in each border region -- Coastal, Northern, Southwest -- for FY 2020?
Method: Group By, Summarize
A: 
```{r}

#cluster these sectors into specific groups using mutate
#puts this in a new column called border_region

#when the sector is in that list of Miami, NoLa, Ramey, put it in the Coastal Border column

#group_by and summarize creates a pivot table, run to see

bp_fy2020 |>
  mutate(
    border_region = case_when(
      sector %in% c("Miami", "New Orleans", "Ramey") ~ "Coastal Border",
      
      sector %in% c(
        "Blaine", "Buffalo", "Detroit", "Grand Forks",
        "Havre", "Houlton", "Spokane", "Swanton"
      ) ~ "Northern Border",
      
      sector %in% c(
        "Big Bend", "Del Rio", "El Centro", "El Paso",
        "Laredo", "Rio Grande Valley", "San Diego", "Tucson", "Yuma"
      ) ~ "Southwest Border",
      
      TRUE ~ NA_character_
    )
  ) |>
  select(border_region, sector, yearly_total) |>
  group_by(border_region) |>
  summarize(
    total_apprehensions = sum(yearly_total, na.rm = TRUE)
  ) |>
  arrange(desc(total_apprehensions))
  
```

## Your Turn
Develop three questions that you can answer using the data in `bp_fy2020` and write the code to answer it.

```{r}
#Are some regions more popular to enter throughout the year than others?

#deselects fy and yearly total columns from the table
#pivot longer makes the table into many rows from the columns

bp_fy2020 |>
  select(-c(fy, yearly_total, sector)) |>
  pivot_longer(
    cols=c(-sector),
    names_to = "month",
    values_to = "apprehensions"
  ) |>
mutate(
  season = case_when(
    month %in% c("october","november","december") ~ "Q1",
    month %in% c("january","february","march") ~ "Q2",
    month %in% c("april","may","june") ~ "Q3",
    month %in% c("july","august","september") ~ "Q4",
    TRUE ~ NA_character_
  )
) |>
  group_by(season) |>
  summarize(
    total_apprehensions = sum(apprehensions, na.rm = TRUE)
  ) |>
  arrange(desc(total_apprehensions))


```